---
title: "Introduction to catviz: Visualizing Causal Assignment Trees"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to catviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 9,
  fig.height = 6
)
```

## Overview

**catviz** helps you visualize the **Causal Assignment Tree (CAT)** — a hierarchical diagram that maps your panel dataset onto the treatment-timing cells used by modern difference-in-differences estimators.

The CAT makes it easy to:

- Verify treatment timing and never-treated group composition
- Inspect subgroup splits (for DDD / DR-DDD designs)
- Communicate your research design to referees and readers
- Check sample sizes **before** running the estimator

Three designs are supported:

| Design | Description |
|--------|-------------|
| **2×2 DiD** | Single treatment cohort, single pre/post split |
| **CSDiD** | Staggered adoption (Callaway & Sant'Anna 2021) |
| **DDD / DR-DDD** | Staggered adoption × binary subgroup |

---

## Installation

```r
# From CRAN (once released)
install.packages("catviz")

# Development version from GitHub
# devtools::install_github("VictorKilanko/catviz")
```

---

## Quick start — three designs in ten lines

```{r quick_start}
library(catviz)
library(dplyr)

# Simulated hospital panel: states adopt treatment at different years
set.seed(1)
df <- expand.grid(
  hospital_id = paste0("H", 1:30),
  year        = 2015:2022
) %>%
  mutate(
    state = rep(paste0("S", 1:6), each = 5)[match(hospital_id, paste0("H", 1:30))],
    g     = c(rep(2017, 10), rep(2019, 10), rep(Inf, 10))[
              match(hospital_id, paste0("H", 1:30))],
    p     = sample(0:1, n(), replace = TRUE)
  )

# --- CSDID (no subgroup) ---
spec_cs <- cat_spec(df, id = "hospital_id", time = "year", g = "g")
cat_plot_tree(spec_cs)

# --- DDD (with subgroup p) ---
spec_ddd <- cat_spec(df, id = "hospital_id", time = "year", g = "g", subgroup = "p")
cat_plot_tree(spec_ddd)

# --- Grayscale (for B&W publications) ---
cat_plot_tree(spec_ddd, grayscale = TRUE)
```

---

## Step-by-step walkthrough

### 1. Build the specification

`cat_spec()` is the entry point. It attaches internal columns to your data
(`.g`, `.NT`, `.node`, etc.) and stores the variable mapping.

```{r cat_spec}
spec <- cat_spec(
  data     = df,
  id       = "hospital_id",   # unit of analysis
  time     = "year",          # time variable
  g        = "g",             # first treatment year (Inf = never treated)
  subgroup = "p",             # binary subgroup (omit for pure CSDID)
  group_id = "state"          # treatment assigned at state level (optional)
)
```

### 2. Add cohort labels (optional but recommended)

`cat_label()` assigns letters (A, B, C, …) to treatment cohorts in
chronological order. Never-treated units receive the last letter.

```{r cat_label}
spec <- cat_label(spec)

# Inspect the label assignments
spec$data %>%
  distinct(.g, .cohort_letter, .g_pretty) %>%
  arrange(.g)
```

### 3. Count units per node

```{r cat_counts}
cat_counts(spec)
```

Each row corresponds to one CAT node. Nodes encode:

- **`.NT == 1`**: never-treated
- **`t < g`**: pre-treatment observations
- **`t ≥ g`**: post-treatment observations
- **`p = 0 / 1`**: subgroup split (DDD only)

### 4. Plot the tree

```{r cat_plot_tree_ddd, fig.width=10, fig.height=7}
cat_plot_tree(spec, counts = TRUE)
```

#### Grayscale version

```{r cat_plot_tree_gray, fig.width=10, fig.height=7}
cat_plot_tree(spec, counts = TRUE, grayscale = TRUE)
```

#### Save to file

```r
cat_plot_tree(spec, save_plot = "my_cat_tree.png")
# or
p <- cat_plot_tree(spec)
cat_save_png(p, "my_cat_tree.png", width = 14, height = 9, dpi = 400)
```

---

## Calling specific plot functions

`cat_plot_tree()` auto-dispatches, but you can call the underlying
functions directly for fine-grained control:

```r
# Explicitly call CSDiD plot
cat_plot_csdid(spec_cs, counts = TRUE)

# Explicitly call DDD plot
cat_plot_ddd(spec_ddd, counts = TRUE)

# Explicitly call 2×2 DiD plot (single cohort)
spec_did <- cat_spec(df %>% mutate(g = ifelse(g==Inf, Inf, 2017)),
                     id = "hospital_id", time = "year", g = "g")
cat_plot_did(spec_did, counts = TRUE)
```

---

## ATT formulas

```{r att_eq}
# DR-DDD ATT formula
cat_att_equation("drddd")

# CSDiD ATT for the p=1 subgroup
cat_att_equation("csdid", subgroup_value = 1)
```

---

## Covariate balance

```{r balance, eval=FALSE}
df$income <- rnorm(nrow(df))
df$age    <- runif(nrow(df), 30, 70)

bal <- cat_balance_table(spec, covariates = c("income", "age"))
cat_balance_plot(bal)
```

---

## Pretrend diagnostics

```{r diag, eval=FALSE}
df$outcome <- rnorm(nrow(df))
spec_diag  <- cat_spec(df, id = "hospital_id", time = "year", g = "g", subgroup = "p")

# DR-DDD pretrend
result <- cat_diag(spec_diag, outcome = "outcome", method = "drddd")
print(result$plot)

# CSDID parallel-gaps
result2 <- cat_diag(spec_diag, outcome = "outcome", method = "csdid")
print(result2$plot)
```

---

## Summary

| Function | Purpose |
|----------|---------|
| `cat_spec()` | Build the CAT specification |
| `cat_label()` | Assign cohort letters |
| `cat_plot_tree()` | Auto-dispatch unified plot (recommended) |
| `cat_plot_did()` | 2×2 DiD tree |
| `cat_plot_csdid()` | CSDiD tree |
| `cat_plot_ddd()` | DDD / DR-DDD tree |
| `cat_counts()` | Count units per node |
| `cat_balance_table()` | Covariate balance SMDs |
| `cat_balance_plot()` | Love plot |
| `cat_diag()` | Pretrend diagnostic |
| `cat_att_equation()` | ATT formula text/LaTeX |
| `cat_save_png()` | Save plot to PNG |
